Neural networks for solving regression and classification problems

### 2022-11-03 doc/paper.tex
Add more on the paper

\documentclass[twocolumn,10pt,cleanfoot]{asme2ej}

\usepackage{graphicx} %% for loading jpg figures
\usepackage{bm}
\usepackage{nicefrac}
\usepackage{mathtools}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{parskip}
\usepackage{listings}
\usepackage{tablefootnote}
\usepackage{float}
\usepackage{xcolor}
\usepackage{hyperref}

@@

\title{Linear regression for approximating real and synthetic data}

%%% first author
\author{Jonatan H. Hanssen
    \affiliation{
	Bachelor Student, Robotics and Intelligent Systems\\
	Department of Informatics\\
	The faculty of Mathmatics and Natural Sciences\\
    Email: jonatahh@ifi.uio.no
    }
}

\author{Eric E. Reber
    \affiliation{
	Bachelor Student, Robotics and Intelligent Systems\\
	Department of Informatics\\
	The faculty of Mathmatics and Natural Sciences\\
    Email: ericer@ifi.uio.no
    }
}

@@

\section{Abstract}
We used linear regression to create a fit of a synthetic dataset generated by the Franke function, and of a real dataset of elevation data from a terrain in Norway. More specifically, we used Ordinary Least Squares (OLS), Ridge and Lasso regression to explore how the different methods perform on different datasets. On the synthetic dataset, we found that Ridge regression allowed us to increase our model complexity and achieve lower MSE scores than OLS, but that Lasso struggled due to computational constraints. On the real dataset, we found that regularization had little impact, as we did not reach model complexities in which overfitting occured.
% Comparing the Mean Squared Error (MSE) for the different methods, we found that Ridge and Lasso outperform OLS for small datasets and high polynomial degrees, but that OLS, given the optimal polynomial degree for the dataset, yields the lowest MSE. Overall we found that the different forms of linear regression fit the synthetic data well. However, when looking at the real datasets, we found that all forms of linear regression performed relatively poorly. We concluded that linear regression perhaps is not the best way of predicting terrain data with rapid and irregular changes in elevation.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \begin{nomenclature}
% \entry{A}{You may include nomenclature here.}
% \entry{$\alpha$}{There are two arguments for each entry of the nomemclature environment, the symbol and the definition.}
% \end{nomenclature}
%
% The primary text heading is  boldface and flushed left with the left margin.  The spacing between the  text and the heading is two line spaces.

@@

One of the core challenges in machine learning is model selection, because we cannot make a priori assessments about how well the model will fit our dataset. A multitude of factors determine the efficacy of our model, the machine learning method we use, the hyperparameters of said method, and the model complexity we select. An uninformed decision in model selection could result in an under- or overfitting to the data, yielding poor results. Thus, it is important that we rigorously test our tuning such that these factors coincide at an intersection, providing the minimal error in unity.

In this paper, we will provide a detailed inspection of the Ordinary Least Squares (OLS), Ridge and Lasso linear regression methods, their hyperparameters and varying model complexities with the aim of making an informed decision about model selection when fitting to different datasets. The datasets consist of synthetic data created by the Franke function, and a real dataset of terrain in Norway. We inform our decisions by plotting the mean squared error (MSE), our cost function, over different polynomial degrees describing model complexity, over different resampling techniques and over different values of lambda, which is the hyperparameter responsible for a suitable regularization. We will perform a bias-variance decomposition for the different linear regression methods to check for under- and overfitting and plot our implementations of the methods up against popular implementations such as scikit learn to ensure that the behaviour of our models is expected.

First, we will go more into depth about the methods and the theory behind them, along with a brief description of our implementation. We will further continue to the results, alongside a critical discussion of their implications. Finally, we will give a conclusion on the optimal model selection for the different datasets.

@@

This sections covers the theoretic background which underlies the methods used in this paper. We use the following shorthands for our mathematical notation.

\subsubsection{Linear regression}

To describe how linear regression works, let us first describe the problem it aims to solve. Assume we have a dataset, which contains a series of inputs $\{x_i\}$, and corresponding target values $\{y_i\}$. Our assumption is that there is some relationship $y_i = f(x_i) + \epsilon_i$ where $\epsilon_i \sim N(0,\sigma^2)$, and we wish to create a model which approximates this relationship. More specifically, we wish to find a vector $\beta$ which gives us the "best" approximation $\bm{\tilde{y}}$ of $\bm{y}$, by using the parameters in the following equation:

\begin{equation}
X \bm{\beta} = \bm{\tilde{y}}
\label{first}
\end{equation}

$X$ is called the design matrix, and is an $n$ by $p$ matrix created by our input data and our choice of basis functions, where $n$ is the amount of different input states we have and $p$ is the number of features we use for our fit. For our problem, $n$ is the number of distinct $x$ values times the number of distinct $y$ values, and $p$ is based on the polynomial degree we use. The basis functions are the summands of a two variable polynomial of degree $N$. The element $x_{ij}$ is calculated by evaluating the basis function corresponding the coloumn $j$ for the input data corresponding to the row $i$. As we can see from Eq.~\ref{first}, each row of this matrix is scalar multiplied by the vector $\beta$, giving us a linear combination of the values of $X_{i*}$ for each $y_i$ in the vector $\bm{\tilde{y}}$. The vector $\bm{\beta}$ is called the parameter vector, and finding the optimal values $\hat{\bm{\beta}}$ is the problem we wish to solve. But for this, we need to define what an "optimal" prediction $\bm{\tilde{y}}$ is.

\subsubsection{The cost function}

To define the optimal prediction, we use a cost function, and the choice of cost function greatly influences what optimal parameters we get.\footnote{It is here important to keep in mind that the cost function used when fitting the model (training) rarely is the same as the one used for testing our prediction} An intuitive cost function is the mean squared error (MSE). This function simply computes the average distance between our prediction and the target value, squared:

\begin{equation}
	C_{mse}(\tilde{\bm{y}}) = \sum_{i=0}^{n-1}(y_i-\tilde{y_i})^2
\end{equation}

We may now plug in the definition of $\bm{\tilde{y}}$.

\begin{equation}
	C_{mse}(\bm{\beta}) = \sum_{i=0}^{n-1}(y_i-X_{i*}\bm{\beta})^2
\end{equation}

We see that we have a function of our parameters $\bm{\beta}$, which describe how much our prediction deviates from our target values. Our problem now is to optimize this function, that is, find the values of $\beta$ which gives the MSE the lowest value. Luckily for us, it can be proven that the second derivative of this function is always positive \cite[106]{linreg}, which means that we can find the global minimum by solving the equation

\begin{gather}
\frac{\delta}{\delta \beta} \sum_{i=0}^{n-1}(y_i-X_{i*}\bm{\beta})^2 = 0 \\
 - \frac{2}{n} X^T(\bm{y} - X\bm{\beta}) = 0
\end{gather}

Solving this for $\beta$ gives us the so called normal equation, an analytical solution\footnote{This solution requires that $X^TX$ is non-singular, which may not be the case. This is solved by using the singular value decomposition of the design matrix, which we will not go into here} for the optimal $\hat{\beta}$.

% \begin{equation}
\begin{gather}
- \frac{2}{n} X^T(\bm{y} - X\bm{\beta}) = 0 \\
X^T(\bm{y} - X\bm{\beta}) = 0 \\
X^T\bm{y} - X^TX\bm{\beta} = 0 \\
X^T\bm{y} = X^TX\bm{\beta} \\
\bm{\hat{\beta}} = (X^TX)^{-1}X^T\bm{y}
\end{gather}

The process we have just described, with MSE as the cost function, is what is known as Ordinary Least Squares (OLS) regression. Using different cost functions when fitting the model leads to different equations for $\bm{\hat{\beta}}$. This leads us to the two next forms of linear regression we use in this paper. But before this, a...

\subsubsection{Brief statistical intermission}

We must remember that our target values are defined as $y_i = f + \epsilon_i$, that is, they are distorted by a stochastic noise $\epsilon_i \sim N(0,\sigma^2)$. We should therefore take a brief intermission to prove some of the statistical qualities of the method we have just described, so as to reassure ourselves that we are justified in applying linear regression to noisy data. We will start by showing that

\begin{equation}
E[y_i] = \sum_j X_{ij}\beta_j
\end{equation}

The proof is as follows:

$\bm{y}$ is defined as the $f(\bm{x}) + \bm{\epsilon}$. We make an assumption that this can be approximated as a matrix multiplication $X\bm{\beta}$ plus an error vector $\bm{\epsilon}$. This means that each element of the vector
$\bm{y}$ can be expressed as follows:

\begin{equation}
y_i = \sum_j x_{ij}\beta_j + \epsilon_i
\end{equation}

If we take the expectation value of this expression we get the following:

\begin{gather}
E[y_i] = E\left[\sum_j x_{ij}\beta_j + \epsilon_i\right] \\
E[y_i] = E\left[\sum_j x_{ij}\beta_j\right] + E[\epsilon_i]
\end{gather}

However, the elements of $X$ are not stochastic, and neither are the elements of $\beta$ and therefore the first expectation value is simply the sum itself. Furthermore, $\epsilon$ is explicitly defined as a normal distribution $N(0,\sigma^2)$, and will by definition have the expectation value $0$. Therefore, we end up with the final expression:

\begin{equation}
E[y_i] = \sum_j x_{ij}\beta_j = \bm{X}_{i,*}\bm{\beta}
\end{equation}

Next, we show that the variance of $y_i$ is equal to the variance of the noise:

\begin{gather}
var[y_i] = E\left[(y_i - E[y_i])^2\right]  \\
= E\left[y_i^2 - 2E[y_i]y_i + E[y_i]^2\right]
\end{gather}

Distributing the outer expectation value function:

\begin{gather}
var[y_i] = E[y_i^2] - 2E\left[E[y_i]\right]E[y_i] + E\left[E[y]^2\right]  \\
= E[y_i^2] - 2E\left[E[y_i]\right]E[y_i] + E\left[E[y]\right]E\left[E[y]\right]
\end{gather}

The result of calculating the expectation value is non-stochastic. This means that $E[E[X]] = E[X]$. From this it follows that

\begin{gather}
var[y_i] = E[y_i^2] - 2E[y_i]E[y_i] + E[y_i]E[y_i]  \\
= E[y_i^2] - E[y_i]E[y_i]  \\
= E[y_i^2] - E[y_i]^2
\end{gather}

The expectation value in the second summand has been proven to be equal to $\bm{X}_{i,*}\bm{\beta}$ above. We therefore have

\begin{gather}
var[y_i] = E[y_i^2] - (\bm{X}_{i,*}\bm{\beta})^2  \\
= E\left[(X_{i,*}\bm{\beta})^2 + X_{i,*}\bm{\beta}\epsilon_i + \epsilon_i^2\right] - (\bm{X}_{i,*}\bm{\beta})^2  \\
= E\left[(X_{i,*}\bm{\beta})^2\right] + E[X_{i,*}\bm{\beta}\epsilon_i] + E[\epsilon_i^2] - (\bm{X}_{i,*}\bm{\beta})^2
\end{gather}

$X_{i,*}\bm{\beta}$ and $\epsilon_i$ are both scalars. Therefore the expectation value can be written as $E[X_{i,*}\bm{\beta}]E[\epsilon_i]$. However, $E[\epsilon_i]$ is by definition $0$, because $\epsilon_i$ is defined as a normal distribution of mean $0$ and variance $\sigma^2$. Furthermore, the expectation value of the non-stochastic $(X_{i,*}\bm{\beta})^2$ is simply the expression itself. We therefore have

\begin{gather}
var[y_i] = (X_{i,*}\bm{\beta})^2 + E[X_{i,*}\bm{\beta}]\cdot 0 + E[\epsilon_i^2] - (\bm{X}_{i,*}\bm{\beta})^2  \\
= E[\epsilon_i^2]
\end{gather}

We can prove that this is equal to the variance of $\epsilon_i$:

\begin{gather}
E[\epsilon_i^2] = \frac{1}{n} \sum_i \epsilon_i^2 \\
var[\epsilon_i] = \frac{1}{n} \sum_i (\epsilon_i - \bar{\epsilon_i})^2 \\
var[\epsilon_i] = \frac{1}{n} \sum_i (\epsilon_i - 0)^2 \\
var[\epsilon_i] = \frac{1}{n} \sum_i \epsilon_i^2 = E[\epsilon_i^2]
\end{gather}

And of course, we know that the variance of $\epsilon_i$ by definition is $\sigma^2$.

\begin{gather}
var[y_i] = E[\epsilon_i^2] = var[\epsilon_i] = \sigma^2 
\end{gather}

We should also prove some qualities about or parameters. We start by proving that $E[\hat{\beta}] = \beta$

Because we are assuming $(X^{T}X)^{-1}X^{T}$ to be deterministic, $E[(X^{T}X)^{-1}X^{T}] = (X^{T}X)^{-1}X^{T}$. Moreover, we know $E[y] = y$, and we can thus rewrite the equation as follows:

\begin{equation}
E[(X^{T}X)^{-1}X^{T}y] = (X^{T}X)^{-1}X^{T}y = \beta
\end{equation}

Next, we prove that $var(\beta) = \sigma^{2}(X^{T}X)^{-1}$.

$$var(\beta) = var((X^{T}X)^{-1}X^{T}y)$$

Because we are assuming $(X^{T}X)^{-1}X^{T}$ to be deterministic and $y$ to be stochastic, we can rewrite the $var(\beta)$ as following:

\begin{gather}
var(\beta) = (X^{T}X)^{-1}X^{T}var(y)((X^{T}X)^{-1}X^{T})^{T} \\
= (X^{T}X)^{-1}X^{T}var(y)(X^{T})^{T}((X^{T}X)^{-1})^{T}
\end{gather}

Since $(X^{T})^{T} = X$ and $(X^{T}X)^{-1})^{T}$ = $(X^{T}X)^{T})^{-1} = (X^{T}X)^{-1}$

\begin{gather}
= (X^{T}X)^{-1}X^{T}var(y)X(X^{T}X)^{-1}
\end{gather}

We know var(y) = $\sigma^{2}$, and since it is a scalar it is commutative. Thus we may move it freely

\begin{gather}
= \sigma^{2}(X^{T}X)^{-1}X^{T}X(X^{T}X)^{-1} \\
= \sigma^{2}(X^{T}X)^{-1}
\end{gather}

A common problem with OLS is what is known as overfitting (we will discuss this in greater detail in later sections). A symptom of overfitting is that the values in the parameter vector $\bm{\beta}$ grow very large, and vary wildly for small changes in our dataset. To counteract this, we can add to the cost function a term which increases with the values in $\bm{\beta}$. Optimizing the cost function would now not only mean finding the smallest difference between target and predicted values, but also minimizing the absolute value of the parameters. We now define a new cost function

\begin{equation}
	C_{lasso}(\bm{\beta}) = \sum_{i=0}^{n-1}(y_i-X_{i*}\bm{\beta})^2 + \lambda \sum_{j=0}^{p-1} \beta_j^2
\end{equation}

$||\beta||_2^2$ is equal to $\beta^T\beta$ and is simply the norm of the parameter vector, squared. With the new hyperparameter $\lambda \in (0, \inf)$ , we can decide how harshly we should punish large values in our parameter vector. Finding an optimal $\lambda$ is an important step in the process of finding the best parameters for our model.

A similar derivation as done in the previous section gives us the normal equation for Ridge regression:

\begin{equation}
	\bm{\hat{\beta}} = (X^TX + \lambda I)^{-1}X^T\bm{y}
\end{equation}

We see that this equation, with a $\lambda > 0$ avoids the problem of an inversion of a non-singular matrix, and gives us an analytical solution for $\bm{\hat{\beta}}$ yet again.\footnote{We should cherish this analytical solution, as it may well be the last time we see a cost function that can be optimized analytically.} By using ridge regresison, we can control the complexity of our model and effectively hinder overfitting.

Another cost function is one where we add the L1-norm (the sum of the absolute values of the parameters), instead of the L2-norm. The new cost function is as follows.

\begin{equation}
	C_{lasso}(\bm{\beta}) = \sum_{i=0}^{n-1}(y_i-X_{i*}\bm{\beta})^2 + \lambda \sum_{j=0}^{p-1} |\beta_j|
\end{equation}

However, the derivative of the absolute value is not continous, so there is no analytical solution for the optimal parameters with this form of linear regression. We must then solve the problem numerically, which can be done by using Gradient Descent. Like Ridge regression, this form of regression controls the complexity of the model, with the added benefit that Lasso can drive parameters to exactly zero.

\subsubsection{Resampling methods: The bootstrap}

Bootstrap is a resampling method divided into simple steps. First, we reshuffle our dataset with replacement, which will provide us with a new dataset. Note that we will already have split our data into train and test data before we begin the bootstrap, so when we reshuffle, we will only reshuffle the training data. The reshuffled datasets act as our samples.

Secondly, we will perform our linear regression on this new dataset and store the predictions of the test data. Thirdly, we repeat for a given large number of iterations. Since we use all the test predictions to calculate the MSE, bias and variance, we will then according to the central limit theorem \cite{CLT} obtain a normal distribution of these performance metrics. The law of large numbers \cite{LLN} then tells us that the mean sample values of the MSE, bias and variance approach the true value of MSE, bias and variance. This is dependent on our assumption that our data is independently and identically distributed.

These approximations of the true value of MSE, bias and variance will become useful when performing our bias-variance trade-off analysis.

\subsubsection{Cross validation}

Cross validation is another resampling technique, in which we reshuffle our data and then split them into k-folds, where k specifies the number of folds. One fold is used as our test data, and the rest is used as training data. We run our linear regression, then switch up which fold is used for testing until all folds have been used. As in bootstrapping, we will approach the true values of for example the mean squared error for our model. 
Theoretically, because both two resampling techniques use the central limit theorem and the law of large numbers, they should return the same values for the MSE, bias and variance given the same dataset. Though resampling techniques can be computationally expensive, they make up for it by aiding in model evaluation by supplying us with performance metrics obtained from a larger number of samples.

\subsubsection{Bias-variance trade-off}

A crucial part of model evaluation is ensuring that the model is neither underfitted nor overfitted, which we can determine by observing the relationship between the bias and the variance. Bias is defined as $(Bias[\tilde{y}])^2 = (y-E[\tilde{y}])^2$, the difference between the expected value of our prediction and the observed value. Models with high bias are usually underfitted, whilst models with low bias are the opposite. Variance on the other hand is defined as $var[\tilde{f}] = \frac{1}{n}\sum(\tilde{y_i}-E[\tilde{y}])^2$, how far away predictions are from the mean prediction. Models with high variance are usually overfitted to the training data, whilst low variance models are the opposite.

A good rule of thumb for finding a good fit is looking at the intercept between bias and variance. This is because the cost function can be decomposed into these two elements. The proof for this is as follows:

Given data $\bm{y} = f(\bm{x}) + \bm{\epsilon}$ and a OLS model $\bm{\tilde{y}} = X\bm{\beta}$, we define
the Mean Squared Error as follows

\begin{gather}
C(X,\beta) = \frac{1}{n}\sum_{i=0}^{n-1}(y_i - \tilde{y_i})^2
\end{gather}

This is simply the expectation value of $(\bm{y} - \bm{\tilde{y}})^2$

\begin{gather}
C(X,\beta) = E\left[(\bm{y} - \bm{\tilde{y}})^2\right]
\end{gather}

To derive the bias variance decomposition, we now add $E[\bm{\tilde{y}}] - E[\bm{\tilde{y}}]$ 
as a sneaky trick. We will also replace $\bm{y}$ with our definition $\bm{y} = \bm{f} + \bm{\epsilon}$

\begin{gather}
C(X,\beta) = E\left[(\bm{f} + \bm{\epsilon} - E[\bm{\tilde{y}}] + E[\bm{\tilde{y}}] - \bm{\tilde{y}})^2\right]
\end{gather}
%%%%%%%%%%%%%%%%%5
\begin{multline}
= E[(\bm{f} + \bm{\epsilon} - E[\bm{\tilde{y}}])^2 + 2[(\bm{f} + \bm{\epsilon} - E[\bm{\tilde{y}}])(E[\bm{\tilde{y}}] \\
	- \bm{\tilde{y}})] +(E[\bm{\tilde{y}}] - \bm{\tilde{y}})^2]
\end{multline}

\begin{multline}
= E[(\bm{f} + \bm{\epsilon} - E[\bm{\tilde{y}}])^2] + \\
	2E[(\bm{f} + \bm{\epsilon} - E[\bm{\tilde{y}}])(E[\bm{\tilde{y}}] - \bm{\tilde{y}})] + E[(E[\bm{\tilde{y}}] - \bm{\tilde{y}})^2]
\end{multline}

Let us focus our attention on the middle term first.

\begin{gather}
2E\left[(\bm{f} + \bm{\epsilon} + E[\bm{\tilde{y}}])(E[\bm{\tilde{y}}] - \bm{\tilde{y}})\right] 
\end{gather}

Expanding the multiplication:

\begin{gather}
2E\left[\bm{f}E[\bm{\tilde{y}}] + \bm{\epsilon}E[\bm{\tilde{y}}] - E[\bm{\tilde{y}}]E[\bm{\tilde{y}}] - \bm{f}\bm{\tilde{y}} - \bm{\epsilon}\bm{\tilde{y}} + E[\bm{\tilde{y}}]\bm{\tilde{y}}\right] 
\end{gather}

Distributing the expectation value:

\begin{gather}
2\left[E[\bm{f}E[\bm{\tilde{y}}]] + E[\bm{\epsilon}E[\bm{\tilde{y}}]] - E[E[\bm{\tilde{y}}]E[\bm{\tilde{y}}]] - E[\bm{f}\bm{\tilde{y}}] - E[\bm{\epsilon}\bm{\tilde{y}}] + E[E[\bm{\tilde{y}}]\bm{\tilde{y}}]\right] 
\end{gather}

We know that $\bm{\tilde{y}}$ and $\bm{f}$ are non-stochastic, and we know that the expectation value of $\bm{\epsilon}$ is $0$. We also know that 
$\bm{\epsilon}$ and $\bm{\tilde{y}}$ are independent, which means we can write $E[\bm{\epsilon}\bm{\tilde{y}}]$ as $E[\bm{\epsilon}]E[\bm{\tilde{y}}]$ We therefore have

\begin{gather}
2\left[\bm{f}\bm{\tilde{y}} + 0 - \bm{\tilde{y}}\bm{\tilde{y}} - \bm{f}\bm{\tilde{y}} - 0 + \bm{\tilde{y}}\bm{\tilde{y}}\right] = 0 
\end{gather}

We then have

\begin{gather}
C(X,\beta) = E\left[(\bm{f} + \bm{\epsilon} - E[\bm{\tilde{y}}])^2\right] + E\left[(E[\bm{\tilde{y}}] - \bm{\tilde{y}})^2\right] 
\end{gather}

The second term is by definition the variance of model.

\begin{gather}
C(X,\beta) = E\left[(\bm{f} + \bm{\epsilon} - E[\bm{\tilde{y}}])^2\right] + var[\bm{\tilde{f}}] 
\end{gather}

And now the final term

\begin{gather}
E\left[(\bm{f} + \bm{\epsilon} - E[\bm{\tilde{y}}])^2\right]
\end{gather}

\begin{multline}
= E[\bm{f}^2 + \bm{f}\bm{\epsilon} - \bm{f}E[\bm{\tilde{y}}] + \bm{f\epsilon} + \bm{\epsilon}^2 - \\ 
	\bm{\epsilon}E[\bm{\tilde{y}}] - \bm{f}E[\bm{\tilde{y}}] - \bm{\epsilon}E[\bm{\tilde{y}}] + E[\bm{\tilde{y}}]^2]
\end{multline}

After distributing the expectation value we get the following, keeping in mind that $E[\bm{f}] = E[\bm{y} - \bm{\epsilon}] = E[\bm{y}] - E[\bm{\epsilon}] = \bm{y}$

\begin{gather}
\bm{y}^2 - 2\bm{y}\bm{\tilde{y}} + E[\bm{\tilde{y}}]^2 + \sigma^2 \\
(\bm{y} + E[\bm{\tilde{y}}]^2 + \sigma^2 
\end{gather}

We see that the quadratic form is equal to $bias[\bm{\tilde{y}}]^2$. We can then conclude that the cost function can be rewritten as follows:

\begin{gather}
C(X,\beta) = bias[\bm{\tilde{y}}]^2 + var[\bm{\tilde{f}}] + \sigma^2 
\end{gather}

>>>>

\documentclass[onecolumn,10pt,cleanfoot]{asme2ej}

\usepackage{graphicx} %% for loading jpg figures
\usepackage{bm}
\usepackage{nicefrac}
\usepackage{mathtools}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{parskip}
\usepackage{listings}
\usepackage{tablefootnote}
\usepackage{float}
\usepackage{xcolor}

@@

\title{Neural networks for solving regression and classification problems}

%%% first author
\author{Jonatan H. Hanssen
    \affiliation{
	Bachelor Student, Robotics and \\
	Intelligent Systems\\ \\[-10pt]
	Department of Informatics\\ \\[-10pt]
	The faculty of Mathematics and \\
	Natural Sciences\\ \\[-10pt]
    Email: jonatahh@ifi.uio.no
    }
}

\author{Eric E. Reber
    \affiliation{
	Bachelor Student, Robotics and \\
	Intelligent Systems\\ \\[-10pt]
	Department of Informatics\\ \\[-10pt]
	The faculty of Mathematics and \\
	Natural Sciences\\ \\[-10pt]
    Email: ericer@ifi.uio.no
    }
}

\author{Gregor Kajda
    \affiliation{
	Bachelor Student, Robotics and \\
	Intelligent Systems\\ \\[-10pt]
	Department of Informatics\\ \\[-10pt]
	The faculty of Mathematics and \\
	Natural Sciences\\ \\[-10pt]
    Email: grzegork@ifi.uio.no
    }
}

@@

\section{Abstract}

We developed a Feed Forward Neural Network (FFNN) and applied it to many common datasets used for benchmarking machine learning algorithms. Furthermore, we compared our results to those attained by linear and logistic regression. To optimize our network, we explored the use of different schedulers in our gradient descent method, and compared their performance against eachother. Overall, we found that our FFNN performed well in all classification tasks we tested, attaining test accuracies between $95$ and $100$ percent for all classification problem. However, we also found that simple logistic regression performed similarly, with a much lower computational cost. For regression on terrain data, we found our neural net lacking, not being able to beat our polynomial fit from our previous project, and only attaining an MSE of 1 BILLION, which is 2000 BILLION lower than our best polynomial fit of the Franke Function. In our gradient descent, we found that SCHEDULER performed best, being 1 BILLION percent better than the average of all other methods.

@@

% \begin{nomenclature}
% \entry{A}{You may include nomenclature here.}
% \entry{$\alpha$}{There are two arguments for each entry of the nomemclature environment, the symbol and the definition.}
% \end{nomenclature}
%
% The primary text heading is  boldface and flushed left with the left margin.  The spacing between the  text and the heading is two line spaces.

@@

Neural Networks are all the rage these days, everyone wants to have a neural network in their home. Ive even seen people install neural networks in their dogs.

@@

This sections covers the theoretic background which underlies the methods used in this paper.

\subsubsection{Logistic Regression}

As part of the research conducted in this project, we introduce a new type of regression model, known as Logistic regression. This is a powerful, yet simple algorithm used in machine learning for the purpose of approximating a probability associated with the occurrence of a specific event given measured data. The fundamental idea behind logistic regression is based on the exact approach used in Linear Regression, meaning that we wish to describe a relationship between dependent and explanatory variables in terms of a set of estimated parameters. These methods differ however in how they resolve the said matter, with Linear regression modeling a set of parameters by solving a set of linear equations, whilst Logistic regression uses the gradient descent algorithm and the sigmoid function for this purpose.

Logistic regression is thus just plain linear regression where the weighted sum of the LHS of the linear equation - with randomly initialized parameters - is fed into an activation function. This in turn allows for the computation of probability related to an event taking place. Rounding of said probability leads to labeling of each input as belonging to one of two classes, meaning that each object fed into the model will be classified as being of class 0 or 1. To begin with however, the probability output by the activation function will be incorrect for majority of inputs, due to the random nature of the initialized parameters. While this may seem to be an issue, the weights can be adjusted to give better classification results through the application of the gradient descent coupled with an adjustable variable called learning rate. As we also shall see, the implementation of an activation function at the output gives a significant advantage over the classic Linear regression in terms of classification power.

\subsubsection{Gradient Descent}

In our previous project, we have primarily dealt with problems that have an analytical solution. This means that finding the optimal solution in these cases is a deterministic process which does not require fine tuning of parameters involved. However, for the methods which are applied to problems in this research - such as Logistic Regression - no analytical solutions exist, thus we will have to turn to numerical techniques to find the ‘ideal’ solution. Our method of choice in this case is gradient descent, an optimization technique which reduces the residuals by utilizing the idea of moving downhill a convex function until a globally minimal scalar value is reached. The gradient itself is nothing more than a first order derivative of a given function, which points in the direction of the greatest change.

The application of gradient descent to our problems as means of finding the ideal solution implies that the parameters be initialized randomly. The most common approach is to set each parameter equal to an arbitrary number ranging between 0 and 1, and iteratively update the parameters by calculating the gradient at each given point in the residual space \cite[preface p. x]{dds}.  For any one-layer model, the gradient function used to tune the weights will take the form of the product of the partial derivative of the cost function with respect to the activation function, and the partial derivative of the activation function with respect to the weights:

\begin{equation}
\frac{\partial{\cal C}}{\partial w_{jk}}  = \frac{\partial C}{\partial a_{j}}\frac{\partial a_j}{\partial w_{jk}}
\end{equation}

Although gradient decent is an excellent algorithm for finding optimal solutions, it can also become quite computationally expensive for problems where large numbers of datapoints are available. However, this issue may be effectively countered by applying a variant of standard gradient approach known as Stochastic Gradient Decent or simply SGD. The SGD is based on randomly choosing one or a small number of instances from the dataset to update the parameters, hence the name stochastic. While it may seem like a poor idea to approximate the gradient instead of using the true gradient, the SGD has proven to give a large computational advantage and will often allow the model to converge into a global minima faster than GD. Furthermore, the stochastic nature of this method can allow SGD to avoid getting stuck in local minima \cite[46]{sr}.

\subsubsection{Schedulers}

Whilst SGD itself gives a significant boost to the learning pace of machine learning models, it is possible to improve the algorithm further by applying a special type of optimization method called scheduling. These methods work by adaptively adjusting the learning rate, thus allowing us to avoid local minima and converge even faster. One of the simpler methods involves adding a fraction of the previous change to the current value used to update the weights, thus keeping a sort of memory of the previous iterations. This is called momentum, due to the physical analogue of a moving particle having momentum. Other more complex methods include keeping running averages of the first and second moments of the gradient, and using this to inform the direction of change in the current update.

\subsubsection{Feedforward Neural Networks}

For some problems, like approximating a simple function like the Franke Function in project 1, it is sufficient to choose a set of basis functions and find their optimal linear combination. Given that our problem can be reasonably approximated by this basis, we can achieve good results. However, by limiting ourselves to linear functions, we are possibly limiting our model, as our dataset may well be better approximated by a different basis, our perhaps not even by a linear combination of functions at all. Because of this, linear regression is not able to understand the interaction between to arbitrary input variables \cite[165]{gbc}. A better approach would be to simply feed our model our features directly, and hope that it may somehow learn the relationship organically.

To learn any arbitrary relationship between our features may at first seem like an intractable problem, but we are actually able to achieve this with relatively small changes to our linear regression methods. Instead of having our prediction vector be a linear combination of our input vector, we use the linear combinations of the inputs to create an intermediate vector, and apply an activation function to this vector. The resulting values from this function are either passed to the output as a linear combination again, or sent to another intermediate vector. By having the activation function be a nonlinear function, such as the sigmoid function, we are able to approximate any continous function to arbitrary accuracy \cite[230]{cmb}. By introducing these intermediate vectors, we have created a layered structure, where information is fed forward from layer to layer. Because of this, we call this model a Multilayer Perceptron or a Feedforward Neural Network.

Mathematically, each node on a layer is a linear combination of all the nodes of the previous layer, fed into the activation function for the current layer.

\begin{equation}
a_i^l = f^l(z_i^l) = f^l\left(\sum_{j=1}^{N_{l-1}} w_{ij}^l a_j^{l-1} + b_i^l\right).
\end{equation}

% For an FFNN with an arbitrary amount of hidden layers, the output layer is thus defined as
%
% \begin{equation}
% a^{l+1}_i = f^{l+1}\left[\!\sum_{j=1}^{N_l} w_{ij}^3 f^l\left(\sum_{k=1}^{N_{l-1}}w_{jk}^{l-1}\left(\dots f^1\left(\sum_{n=1}^{N_0} w_{mn}^1 x_n+ b_m^1\right)\dots\right)+b_k^2\right)+b_1^3\right].
% \end{equation}
%
% \begin{equation}
% 	a^{L}_i = f^{L}\left[\!\sum_{j=1}^{N_{L-1}} w_{ij}^3 f^{L-1}\left(\sum_{k=1}^{N_{L-2}}w_{jk}^{L-2}\left(\dots f^1\left(\sum_{n=1}^{N_0} w_{mn}^1 x_n+ b_m^1\right)\dots\right)+b_k^2\right)+b_1^3\right].
% \end{equation}

\subsubsection{Activation functions}

As we have seen, the behaviour of our neural network is defined by the amount of hidden layers and the amount of nodes in each one (our architecture), but also by our choice of activation functions. For the output layer, the activation is usually easily decided based on the problem (identity for regression problems, sigmoid for binary classification, SoftMax for multiclass). For the hidden layers, the choice of activation function is often decided through a process of trial and error, as this is an area without definitive guiding theoretical principles \cite[188]{gbc}. Below are the definitions of some of the most common activation functions.

The sigmoid is defined as follows:

\begin{equation}
y = \frac{1}{1 + exp(x)}
\end{equation}

As we can see from the equation, this function has a range of $y \in (0,1)$, and for high values of x, the rate of change is very small. This saturation is a problem for gradient descent methods, and especially for backpropagation, as the repeated multiplication of very small gradients can lead to weights barely being updated and training slowing down (or even stopping completely). Because of this saturation, the sigmoid is generally not used as an activation function for the hidden layers \cite[191]{gbc}.

Another activation function is the Rectified Linear Unit, and its cousin the Leaky Rectified Linear Unit:

\begin{equation}
RELU(x) = \left\{\begin{array}{cc} 0 & \; \; if \; \; x < 0, \\  x & \; \; if \; \; x \ge 0.\end{array}\right.
\end{equation}

\begin{equation}
LRELU(x) = \left\{\begin{array}{cc} -\alpha \cdot x & \; \; if \; \; x < 0, \\  x & \; \; if \; \; x \ge 0.\end{array}\right.
\end{equation}

These do not suffer the same saturation problems displayed by the sigmoid function, and as such are often the default choice of activation for the hidden layers \cite[188]{gbc}.

\subsubsection{Backpropagation}

Like logistic regression, the optimal weights for each layer in an FFNN can not be found analytically, and we have to turn to gradient descent. By repeated use of the chain rule, we are able to calculate the gradient for each layer iteratively, using the gradient of the layer after it and starting at the output. This is simply the reverse mode of automatic differentiation, and is computationally preferred compared to forward mode, where we start at the input layer \cite[416]{sr}. The equations for calculating the gradients in an FFNN, which we will not derive here, are as follows \cite{morten}:

First we calculate the so called delta terms for all layers, starting with the output layer:

\begin{equation}
y_i^l = f^l(u_i^l) = f^l\left(\sum_{j=1}^{N_{l-1}} w_{ij}^l y_j^{l-1} + b_i^l\right).
\end{equation}

We use this term in an iterative process, working backwards through the network with the following equation:

\begin{equation}
\delta_j^l = \sum_k \delta_k^{l+1}w_{kj}^{l+1}f'(z_j^l).
\end{equation}

Using these delta terms, we can calculate the gradients for the weights in each layer:

\begin{gather}
w_{jk}^l\leftarrow  = w_{jk}^l- \eta \delta_j^la_k^{l-1} \\
b_j^l \leftarrow b_j^l-\eta \frac{\partial {\cal C}}{\partial b_j^l}=b_j^l-\eta \delta_j^l
\end{gather}
